Detailed disease progression of 213 patients hospitalized with Covid-19 in the Czech Republic: An exploratory analysis

### 2020-11-30 manuscript/manuscript.Rmd
Minor tweaks based on final feedback

The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020, led to over 46 million cases and over 1.1 million deaths. Proposed treatments include antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR]. Currently, the only treatment that exhibited positive effects in randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58]. Remdesivir, previously designed and approved for Ebola, SARS and MERS, is currently the only antiviral against SARS-CoV-2 approved by the FDA. However, during spring and summer it was available only in limited amounts via compassionate use programme. Randomized clinical trials on remdesivir have provided  conflicting evidence [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/KB5UDZH3]. Despite initial promise, hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], and as both pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR]. Further, no improvement could be discerned when adding azithromycin to hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, Lopinavir/Ritonavir also did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].

@@

High IL-6, D-dimer values were observed to be associated with outcome and disease severity [@http://zotero.org/users/5567156/items/K3N3VG2X].
Large study of electronic health records [@http://zotero.org/users/5567156/items/ARTB8HYZ] showed an increase in C-reactive protein in early disease and increase of D-dimer and white blood cell count in later stages of the disease.

@@

Not all patients developed pneumonia or other symptoms of Covid-19. All patients received the standard of care including supplemental oxygen and ventilation, antibiotics for bacterial superinfections. Some patients were not indicated for all treatment modalities (especially mechanical ventilation) based on decision of the attending physician and underlying patient condition. We note that the participating sites were not homogeneous in either patient population or treatment protocols. The choice of treatment was based on the decision of the attending clinician and availability.

@@

```{r modelstates, fig.cap="States used in the competing risk model (a) and in the two hidden Markov model variants (b,c). AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation. In all models the 'Death' and 'Discharged' states are terminal. In the second hidden Markov model (c), the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
knitr::include_graphics(here::here("manuscript","static_figs","model_states.png"), auto_pdf = TRUE)
```

@@

data_table_1 %>% group_by(group) %>% 
  summarise(
    N = n(),
    `Distinct sites` = length(unique(hospital_id)),
    Male = sum_percent(sex == "M"),
    `Age (mean, IQR)` = mean_iqr(age),
    `Admitted for Covid` = sum_percent(admitted_for_covid),
    `Took azithromycin` = sum_percent(ever_az),
    `Took dexamethasone` = sum_percent(ever_dexamethasone),
    `Took favipiravir` = sum_percent(ever_favipiravir),
    `Took remdesivir` = sum_percent(ever_remdesivir),
    `Convalescent plasma` = sum_percent(ever_convalescent_plasma),
    `Ischemic Heart Disease` = sum_percent(ischemic_heart_disease),
    `Takes antihypertensives` = sum_percent(has_hypertension_drugs),	
    `Heart Failure` = sum_percent(heart_failure),
    COPD = sum_percent(COPD),
    Asthma = sum_percent(asthma),
    `Other lung disease` = sum_percent(other_lung_disease),
    Diabetes = sum_percent(diabetes),
    `Renal Disease` = sum_percent(renal_disease),
    `Liver Disease` = sum_percent(liver_disease),
    `Smoking` = sum_percent(smoking),
    `BMI (mean, IQR)` = mean_iqr(BMI),
    `Best supportive care` = sum_percent(!is.na(best_supportive_care_from)),
    Deceased = sum_percent(outcome == "Death"),
    Discharged = sum_percent(outcome == "Discharged"),
    .groups = "drop"
  ) %>%
  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Favipiravir"), ordered = TRUE)) %>%
  arrange(group) %>%
  as.data.frame() %>%
  column_to_rownames("group") %>%
  t() %>% 
  knitr::kable(caption = "Patient characteristics for the overall sample and treatment subgroups. Note that the favipiravir subgroup is not exclusive with either the HCQ or No HCQ group. IQR = interquartile range, COPD = Chronic obstructive pulmonary disease, BMI = body-mass index, Best supportive care = patient was not indicated for progression to a more intensive treatment modality.", booktabs = TRUE)

@@

In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure&nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured. The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients had high CRP upon admission and for many the CRP levels stayed elevated over the whole hospitalization. The limited nature of the data however does not allow for any statistically robust conclusions. We also see that the marker levels were not substantially stratified by study site.  Those patterns should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured. However we believe this kind of patient-level view is useful to appreciate the extent of both between-patient and within-patient variability.

```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required. Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation."}
my_paste <- function(x) {
  paste0(as.integer(x), collapse = "")
}

@@

```{r selectedmarkers, fig.cap = "Values of selected markers over the course of the disease. Each line represents a patient, stratified by the worst breathing support required over the hospitalization. Color indicates study sites. The vertical scale is logarithmic. Ventilated = any form of ventilation, including non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation, CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [10^9/l], IL-6 = Interleukin 6 [ng/l]."}
#data$marker_data %>%
# TODO: use SpO2 in AA as SpO2 native (and vice versa)
plot_marker_data <- function(markers, log_transform, smooth = FALSE, max_patients = NULL, max_day = NULL) {

@@

Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to future hydroxychloroquine treatment:
the score representing the cumulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
other lung diseases, renal disease, or high creatinine 
was associated with a lower chance of taking hydroxychloroquine over the course of the hospitalization 
(the chance was only `r round(100*exp(coef(m.hcq_treatment2)[2]),1)`%, 95% confidence interval 
(`r round(100*exp(confint(m.hcq_treatment2)[2,]),1)`)%,
Chi-square test in the logistic regression model, 
$\chi^2$=`r round(abs(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq'))[2,4],2)`, df=1, 
P=`r round(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq')[2,5],3)`).

@@

We also examined the association between treatments and length of hospital stay for patients that survived. Almost all models cannot discern the sign of the association for all treatments examined (Supplementary Figures `r supplementary_figures_multiverse`). Similarly, we studied the association between D-dimer and Interleukin 6 and outcomes, with unconclusive results as well (Supplementary Figure `r supplementary_figure_markers` )

@@

All prognostic models we tested performed similarly to or notably worse than using age as the only predictor and also worse than originally reported (Figure&nbsp;\@ref(fig:riskscores)). Additionally, some publications did not provide enough detail to unambigously reconstruct how the score and/or outcome was assessed. We thus concur with Wynants et al. [@http://zotero.org/users/5567156/items/TS57HJ3T] that reported prediction scores are at high risk of bias and need additional careful evaluation.

@@

Additionally our experience indicates that a substantial fraction of published prognostic models will perform much worse on new patients than on the datasets they were built for and that external validation is crucial. We suggest that comparing the prognostic models against simple baselines (e.g. decade of age as the single predictor) should be a first step in validation. Further, some of the published scores lack enough information to let others exactly implement the score.

>>>>

The Covid-19 pandemic, caused by severe acute respiratory syndrome coronavirus (SARS-CoV-2) has, as of November 2020, led to over 46 million cases and over 1.1 million deaths. Proposed treatments include antivirals approved for other indications (chloroquine, hydroxychloroquine, lopinavir/ritonavir, remdesivir, favipiravir, umifenovir), azithromycin, corticosteroids, immunoglobulins, tocilizumab and convalescent plasma [@http://zotero.org/users/5567156/items/8XWTUMUI; @http://zotero.org/users/5567156/items/6Q8Z8VJR]. As of this writing, the only treatment that exhibited positive effects in randomized trials is the corticosteroid dexamethasone for severe cases [@http://zotero.org/users/5567156/items/IW6VZ53Q; @http://zotero.org/users/5567156/items/5VAKEW58]. Remdesivir, previously designed and approved for Ebola, SARS and MERS, is currently the only antiviral against SARS-CoV-2 approved by the FDA. However, during spring and summer it was available only in limited amounts via compassionate use programme. Randomized clinical trials on remdesivir have provided  conflicting evidence [@http://zotero.org/users/5567156/items/MFS3IPPU; @http://zotero.org/users/5567156/items/KB5UDZH3]. Despite initial promise, hydroxychloroquine has failed to show benefit for both hospitalized patients [@http://zotero.org/users/5567156/items/9X3UH8ZM; @http://zotero.org/users/5567156/items/KB5UDZH3], as an early treatment [@http://zotero.org/users/5567156/items/G6KZKVMI], as well as pre- and post-exposure prophylaxis [@http://zotero.org/users/5567156/items/7C23YCV5; @http://zotero.org/users/5567156/items/GMR9FX9Z; @http://zotero.org/users/5567156/items/H4CLJNUR]. Further, no improvement could be discerned when adding azithromycin to hydroxychloroquine therapy [@http://zotero.org/users/5567156/items/BQLRZGTQ]. Similarly, Lopinavir/Ritonavir did not show benefit [@http://zotero.org/users/5567156/items/6I6GAVWU]. No benefit was observed for convalescent plasma [@http://zotero.org/users/5567156/items/UEXLDJCB; @http://zotero.org/users/5567156/items/BHGT88US].

@@

High IL-6, D-dimer values were observed to be associated with worse outcome and increased disease severity [@http://zotero.org/users/5567156/items/K3N3VG2X].
Large study of electronic health records [@http://zotero.org/users/5567156/items/ARTB8HYZ] showed an increase in C-reactive protein in early disease and increase of D-dimer and white blood cell count in later stages of the disease.

@@

Not all patients developed pneumonia or other symptoms of Covid-19. All patients received the standard of care including supplemental oxygen and ventilation and antibiotics for bacterial superinfections as determined by the attending physician. Some patients were not indicated for all treatment modalities (especially mechanical ventilation) based on decision of the attending physician and underlying patient condition. We note that the participating sites were not homogeneous in either patient population or treatment protocols. The choice of pharmacological treatment was based on the decision of the attending clinician and its availability.

@@

```{r modelstates, fig.cap="States used in the competing risk model (a) and in the two hidden Markov model variants (b,c). AA = Ambient air, Oxygen = Nasal oxygen, Ventilated = any form of ventilation (non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation). In all models the 'Death' and 'Discharged' states are terminal. In the second hidden Markov model (c), the 'Improving' and 'Worsening' variants of each non-terminal state are not observable - only the breathing support is observed and improving/worsening is inferred from progression of the disease." }
knitr::include_graphics(here::here("manuscript","static_figs","model_states.png"), auto_pdf = TRUE)
```

@@

data_table_1 %>% group_by(group) %>% 
  summarise(
    N = n(),
    `Distinct sites` = length(unique(hospital_id)),
    Male = sum_percent(sex == "M"),
    `Age (mean, IQR)` = mean_iqr(age),
    `Admitted for Covid` = sum_percent(admitted_for_covid),
    `Took azithromycin` = sum_percent(ever_az),
    `Took dexamethasone` = sum_percent(ever_dexamethasone),
    `Took favipiravir` = sum_percent(ever_favipiravir),
    `Took remdesivir` = sum_percent(ever_remdesivir),
    `Convalescent plasma` = sum_percent(ever_convalescent_plasma),
    `Ischemic Heart Disease` = sum_percent(ischemic_heart_disease),
    `Takes antihypertensives` = sum_percent(has_hypertension_drugs),	
    `Heart Failure` = sum_percent(heart_failure),
    COPD = sum_percent(COPD),
    Asthma = sum_percent(asthma),
    `Other lung disease` = sum_percent(other_lung_disease),
    Diabetes = sum_percent(diabetes),
    `Renal Disease` = sum_percent(renal_disease),
    `Liver Disease` = sum_percent(liver_disease),
    `Smoking` = sum_percent(smoking),
    `BMI (mean, IQR)` = mean_iqr(BMI),
    `Best supportive care` = sum_percent(!is.na(best_supportive_care_from)),
    Deceased = sum_percent(outcome == "Death"),
    Discharged = sum_percent(outcome == "Discharged"),
    .groups = "drop"
  ) %>%
  mutate(group = factor(group, levels= c("All", "HCQ", "No HCQ", "Favipiravir"), ordered = TRUE)) %>%
  arrange(group) %>%
  as.data.frame() %>%
  column_to_rownames("group") %>%
  t() %>% 
  knitr::kable(caption = "Patient characteristics for the overall sample and treatment subgroups. Note that the favipiravir subgroup is not exclusive with either the HCQ or No HCQ group. IQR = interquartile range, COPD = Chronic obstructive pulmonary disease, BMI = body-mass index, Best supportive care = patient was not indicated to undergo more intensive treatment modality.", booktabs = TRUE)

@@

In Figure&nbsp;\@ref(fig:progression) we show the overall disease progression for all patients and in Figure&nbsp;\@ref(fig:selectedmarkers) we show the time-course of a subset of the markers we have measured. The data show some interesting patterns: patients with low Interleukin-6 or D-dimer values are overrepresented among patients with better outcomes, most patients had high CRP upon admission and for many the CRP levels stayed elevated over the whole hospitalization. However, the limited nature of the data does not allow for any statistically robust conclusions. We also see that the marker levels were not substantially stratified by study site.  Those patterns should however be interpreted with care due to systematic missingness of the data - in particular, patients that fared worse were probably more likely to have the markers measured. However we believe this kind of patient-level view is useful to appreciate the extent of both between-patient and within-patient variability.

```{r progression, fig.cap="Disease progression for all patients included in the study as determined by breathing support required. Ventilated = any form of ventilation (non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation)."}
my_paste <- function(x) {
  paste0(as.integer(x), collapse = "")
}

@@

```{r selectedmarkers, fig.cap = "Values of selected markers over the course of the disease. Each line represents a patient, stratified by the worst breathing support required over the hospitalization. Color indicates study sites. The vertical scale is logarithmic. Ventilated = any form of ventilation (non-invasive positive-pressure ventilation, mechanical ventilation and extra-corporeal membrane oxygenation), CRP = C-reactive protein [mg/l], D-dimer [ng/ml DDU], Ly = lymphocyte count [10^9/l], IL-6 = Interleukin 6 [ng/l]."}
#data$marker_data %>%
# TODO: use SpO2 in AA as SpO2 native (and vice versa)
plot_marker_data <- function(markers, log_transform, smooth = FALSE, max_patients = NULL, max_day = NULL) {

@@

Taken quantitatively, the comorbidities known upon hospitalization were informative with respect to the future hydroxychloroquine treatment:
the score representing the cumulative presence of ischemic heart disease, hypertension drugs, former heart failure, COPD, 
other lung diseases, renal disease, or high creatinine 
was associated with a lower chance of taking hydroxychloroquine over the course of the hospitalization 
(the chance was only `r round(100*exp(coef(m.hcq_treatment2)[2]),1)`%, 95% confidence interval 
(`r round(100*exp(confint(m.hcq_treatment2)[2,]),1)`)%,
Chi-square test in the logistic regression model, 
$\chi^2$=`r round(abs(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq'))[2,4],2)`, df=1, 
P=`r round(anova(m.hcq_treatment2,m.hcq_treatment0,test='Chisq')[2,5],3)`).

@@

We also examined the association between treatments and length of hospital stay for all the patients that survived. Almost all models cannot discern the sign of the association for all treatments examined (Supplementary Figures `r supplementary_figures_multiverse`). Similarly, we studied the association between D-dimer and Interleukin 6 and outcomes, with unconclusive results as well (Supplementary Figure `r supplementary_figure_markers` )

@@

All prognostic models we tested performed similarly to or notably worse than using age as the only predictor and also worse than originally reported (Figure&nbsp;\@ref(fig:riskscores)). Additionally, some publications did not provide enough detail to unambiguously reconstruct how the score and/or outcome was assessed. We thus concur with Wynants et al. [@http://zotero.org/users/5567156/items/TS57HJ3T] that reported prediction scores are at high risk of bias and need additional careful evaluation.

@@

Additionally, our experience indicates that a substantial fraction of published prognostic models will perform much worse on new patients than on the datasets they were built for and that external validation is crucial. We suggest that comparing the prognostic models against simple baselines (e.g. decade of age as the single predictor) should be a first step in validation. Furthermore, some of the published scores lack enough information to let others implement the score in the same way.
